---
output: html_document
editor_options:
  chunk_output_type: console
---

# Explainable AI

```{r}
#| echo: false
#| include: false
#| results: false
reticulate::use_virtualenv("r-keras")
library(tensorflow)
library(raster)
library(sp)
tf
tf$abs(3.)
```

The goal of explainable AI (xAI, aka interpretable machine learning) is to explain **why** a fitted machine learning model makes certain predictions. A typical example is to understand how important different variables are for predictions. The incentives for doing so range from a better technical understanding of the models over understanding which data is important for improving predictions to questions of fairness and discrimination (e.g. to understand if an algorithm uses skin color to make a decision).

## A Practical Example

In this lecture we will work with an African Elephant occurrence dataset.

We will fit a random forest and use the iml package for xAI, see <a href="https://christophm.github.io/interpretable-ml-book/" target="_blank" rel="noopener">https://christophm.github.io/interpretable-ml-book/</a>.

```{r chunk_chapter6_0, eval=FALSE}
library(iml)
library(ranger) # different random Forest package!
library(EcoData)
library(cito)
set.seed(123)


data = EcoData::elephant$occurenceData
head(data)
?EcoData::elephant

```

Meaning of the bioclim variables:

| Bioclim variable | Meaning |
|--------------------------------------|----------------------------------|
| bio1 | Annual Mean Temperature |
| bio2 | Mean Diurnal Range (Mean of monthly (max temp - min temp)) |
| bio3 | Isothermality (BIO2/BIO7) (×100) |
| bio4 | Temperature Seasonality (standard deviation ×100) |
| bio5 | Max Temperature of Warmest Month |
| bio6 | Min Temperature of Coldest Month |
| bio7 | Temperature Annual Range (BIO5-BIO6) |
| bio8 | Mean Temperature of Wettest Quarter |
| bio9 | Mean Temperature of Driest Quarter |
| bio10 | Mean Temperature of Warmest Quarter |
| bio11 | Mean Temperature of Coldest Quarter |
| bio12 | Annual Precipitation |
| bio13 | Precipitation of Wettest Month |
| bio14 | Precipitation of Driest Month |
| bio15 | Precipitation Seasonality (Coefficient of Variation) |
| bio16 | Precipitation of Wettest Quarter |
| bio17 | Precipitation of Driest Quarter |
| bio18 | Precipitation of Warmest Quarter |
| bio19 | Precipitation of Coldest Quarter |

```{r, eval=FALSE}
rf = ranger(as.factor(Presence) ~ ., data = data, probability = TRUE)
```

::: callout-note
## ranger Package

The ranger package provides an alternative implementation of the random forest algorithm. The implementation in the ranger package is one of the fastest available, which is especially important for explainable AI. Most xAI tools require hundreds of predictions (e.g. feature importance permutes each feature n times to calculate the performance drop), so a fast implementation of the ML algorithm is crucial.

**Important**:

For binary classification tasks, it is critical to change the response variable to a factor before fitting ranger! Otherwise, ranger will use the first value in the response variable as "Class 1"!
:::

The cito package has quite extensive xAI functionalities. However, ranger, as most other machine learning packages, has no extensive xAI functionalities. Thus, to do xAI with ranger, we have to use a generic xAI package that can handle almost all machine learning models.

When we want to use such a generic package, we first have to create a predictor object, that holds the model and the data. The `iml` package uses R6 classes, that means new objects can be created by calling `Predictor$new()`. (Do not worry if you do not know what R6 classes are, just use the command.)

To make the xAI tools available to many different packages/algorithms, the `iml` package expects that the ML algorithm specific predict method to be wrapped in a generic predict function in the form of `function(model, newdata) predict(model, newdata)` and the function wrapper should return a vector of predictions:

```{r chunk_chapter6_1, eval=FALSE}

predict_wrapper = function(model, newdata) predict(model, data=newdata)$predictions[,2]

predictor = Predictor$new(rf, data = data[,-1], y = data[,1], predict.function = predict_wrapper)
predictor$task = "classif" # set task to classification
# "Predictor" is an object generator.
```

## Feature/Permutation Importance

Feature importance should not be confused with random forest variable importance, although they are related. It tells us how important each variable is for prediction, can be computed for all machine learning models, and is based on a permutation approach (see the book):

```{r chunk_chapter6_2, eval=FALSE}
imp = FeatureImp$new(predictor, loss = "ce")
plot(imp)
```

bio9 (Precipitation of the wettest Quarter) is the most important variable.

## Partial Dependencies

Partial dependencies are similar to `allEffects` plots for normal regressions. The idea is to visualize "marginal effects" of predictors (with the "feature" argument we specify the variable we want to visualize):

```{r chunk_chapter6_3, eval=FALSE}
eff = FeatureEffect$new(predictor, feature = "bio9", method = "pdp",
                        grid.size = 30)
plot(eff)
```

One disadvantage of partial dependencies is that they are sensitive to correlated predictors. Accumulated local effects can be used for accounting for correlation of predictors.

## Accumulated Local Effects

Accumulated local effects (ALE) are basically partial dependencies plots but try to correct for correlations between predictors.

```{r chunk_chapter6_5, eval=FALSE}
ale = FeatureEffect$new(predictor, feature = "bio9", method = "ale")
ale$plot()
```

If there is no collinearity, you shouldn't see much difference between partial dependencies and ALE plots.

## Friedman's H-statistic

The H-statistic can be used to find interactions between predictors. However, again, keep in mind that the H-statistic is sensible to correlation between predictors:

```{r chunk_chapter6_6, message=FALSE, warning=FALSE, eval=FALSE}
interact = Interaction$new(predictor, "bio9",grid.size = 5L)
plot(interact)
```

## Global Explainer - Simplifying the Machine Learning Model

Another idea is simplifying the machine learning model with another simpler model such as a decision tree. We create predictions with the machine learning model for a lot of different input values and then we fit a decision tree on these predictions. We can then interpret the easier model.

```{r chunk_chapter6_7, results='hide', message=FALSE, warning=FALSE, eval=FALSE}
library(partykit)

tree = TreeSurrogate$new(predictor, maxdepth = 2)
plot(tree$tree)
```

## Local Explainer - LIME Explaining Single Instances (observations)

The global approach is to simplify the entire machine learning-black-box model via a simpler model, which is then interpretable.

However, sometimes we are only interested in understanding how single predictions are generated. The LIME (Local interpretable model-agnostic explanations) approach explores the feature space around one observation and based on this locally fits a simpler model (e.g. a linear model):

```{r chunk_chapter6_8, message=FALSE, warning=FALSE, eval=FALSE}
lime.explain = LocalModel$new(predictor, x.interest = data[1,-1])
lime.explain$results
plot(lime.explain)
```

## Local Explainer - Shapley

The Shapley method computes the so called Shapley value, feature contributions for single predictions, and is based on an approach from cooperative game theory. The idea is that each feature value of the instance is a "player" in a game, where the prediction is the reward. The Shapley value tells us how to fairly distribute the reward among the features.

```{r chunk_chapter6_9, eval=FALSE}
shapley = Shapley$new(predictor, x.interest = data[1,-1])
shapley$plot()
```

## Uncertainties - the bootstrap

Standard xAI method do not provide reliable uncertainties on the fitted curves. If you want uncertainties or p-values, the most common method is the bootstrap.

In a bootstrap, is instead of splitting up the data in test / validation, we sample from the data with replacement and fit the models repeatedly. The idea is to get an estimate about the variability we would expect if we created another dataset of the same size.

```{r, eval=FALSE}
k = 10 # bootstrap samples
n = nrow(data)
error = rep(NA, k)

for(i in 1:k){
  bootSample = sample.int(n, n, replace = TRUE)
  rf = ranger(as.factor(Presence) ~ ., data = data[bootSample,], probability = TRUE)
  error[i] = rf$prediction.error
}

hist(error, main = "uncertainty of in-sample error")
```

Note that the distinction between bootstrap and validation / cross-validation is as follows:

-   Validation / cross-validation estimates out-of-sample predictive error
-   Bootstrap estimates uncertainty / confidence interval on all model outputs (could be prediction and inference).

## Exercises

::: callout-warning
#### xAI in cito

Data preparation

```{r, eval=FALSE}
library(iml)
library(cito)
library(EcoData)
library(cito)


data = EcoData::elephant$occurenceData
head(data)
?EcoData::elephant

# we will subsample data (absences) to reduce runtime
data_sub = data[sample.int(nrow(data), 500),]

```

Cito includes serveral xAI methods directly out of the box

```{r, eval=FALSE}
model = dnn(Presence~., data = data_sub, batchsize = 200L,loss = "binomial", verbose = FALSE, lr = 0.15, epochs = 300)
```

Try the following commands:

-   `summary(dnn, n_permute = 10)`
-   `PDP(dnn)`
-   `ALE(dnn)`

Moreover, try to refit the model with the option `bootstrap = 5`. This may take a short while. Observe how the xAI options change.
:::

`r hide("Click here to see the solution for cito")`

```{r, eval=FALSE}
model = dnn(Presence~., data = data_sub, batchsize = 200L, bootstrap = 5L, loss = "binomial", verbose = FALSE, lr = 0.15, epochs = 300)
```

```{r, eval=FALSE}
summary(model, n_permute = 10L)
```

```{r, eval=FALSE}
PDP(model)
ALE(model)
```

`r unhide()`

::: callout-warning
#### Question

Use the Titanic_ml dataset and fit a random forest, dnn or a BRT using xgboost. Explore / interpret the fitted model using `iml` (see also the book: <a href="https://christophm.github.io/interpretable-ml-book/" target="_blank" rel="noopener">https://christophm.github.io/interpretable-ml-book/</a>).

Tip:

If you use iml, you need to provide a proper prediction function wrapper:

```{r, eval=FALSE}
# random Forest (ranger), regression:
predict_wrapper = function(model, newdata) predict(model, data=newdata)$predictions

# random Forest (ranger), classification:
predict_wrapper = function(model, newdata) predict(model, data=newdata)$predictions[,2]

# xgboost:
predict_wrapper = function(model, newdata) predict(model, as.matrix(newdata))
```

Prepare the data

```{r, eval=FALSE}
library(EcoData)
library(dplyr)
library(missRanger) # for imputation


data = titanic_ml

# feature selection
data = data %>% select(survived, sex, age, fare, pclass)# play around with the features

# imputation - remove response variable!
head(data)
data_imputed = data
data_imputed[,-1] = missRanger(data_imputed[,-1])
summary(data_imputed)

data_imputed = data_imputed %>% 
  mutate(age = (age - mean(age))/sd(age), fare = (fare - mean(fare))/sd(fare),
         sex = as.integer(sex), pclass = as.integer(pclass))

data_obs = data_imputed[!is.na(data_imputed$survived), ]
data_new = data_imputed[is.na(data_imputed$survived), ]
```
:::

`r hide("Click here to see the solution for RF")`

```{r , message=FALSE, warning=FALSE, eval=FALSE}
library(ranger)
library("iml")
set.seed(1234)
data_obs$survived = as.factor(data_obs$survived)

rf = ranger(survived ~ ., data = data_obs, importance = "impurity", probability = TRUE)

# For submission:
#write.csv(data.frame(y=predict(rf, data_new)$predictions[,2]), file = "wine_RF.csv")

# Standard depiction of importance:
ranger::importance(rf)

# Setup wrapper
predict_wrapper = function(model, newdata) predict(model, data=newdata)$predictions[,2]


# IML:
predictor = Predictor$new(
    rf, data = data_obs[,which(names(data_obs) != "survived")], y = as.integer(data_obs$survived)-1,
    predict.function = predict_wrapper
    )

# Mind: This is stochastical!
importance = FeatureImp$new(predictor, loss = "logLoss")

plot(importance)

# Comparison between standard importance and IML importance:
importanceRf = names(rf$variable.importance)[order(rf$variable.importance, decreasing = TRUE)]
importanceIML = importance$results[1]
comparison = cbind(importanceIML, importanceRf)
colnames(comparison) = c("IML", "RF")
as.matrix(comparison)
```

Mind that feature importance, and the random forest's variable importance are related but not equal! Variable importance is a measure for determining importance while creating the forest (i.e. for fitting). Feature importance is a measure for how important a variable is for prediction.

Maybe you want to see other explanation methods as well. Surely you can use the other techniques of this section on your own.

`r unhide()`

`r hide("Click here to see the solution for xgboost")`

```{r chunk_chapter6_task_00, message=FALSE, warning=FALSE, eval=FALSE}
library(xgboost)
library("iml")
set.seed(1234)


data_xg = xgb.DMatrix(
  data = as.matrix(data_obs[,which(names(data_obs) != "survived")]),
  label = as.integer(data_obs$survived)-1
)
brt = xgboost(data_xg, nrounds = 24, objective = "reg:logistic")


# For submission:
#write.csv(round(predict(brt, data_new)), file = "wine_RF.csv")

# Standard depiction of importance:
xgboost::xgb.importance(model = brt)

# Setup wrapper
predict_wrapper = function(model, newdata) predict(model, as.matrix(newdata))


# IML:
predictor = Predictor$new(
    brt, data = data_obs[,which(names(data_obs) != "survived")], y = as.integer(data_obs$survived)-1,
    predict.function = predict_wrapper
    )

# Mind: This is stochastical!
importance = FeatureImp$new(predictor, loss = "logLoss")

plot(importance)

```

`r unhide()`

`r hide("Click here to see the solution for cito")`

```{r chunk_chapter6_task_0, message=FALSE, warning=FALSE, eval=FALSE}
library(cito)
data_obs$survived = as.integer(data_obs$survived) - 1
nn = dnn(survived~., data = data_obs, loss = "binomial", lr= 0.03, epochs = 300)

summary(nn)
```

`r unhide()`

<!-- ::: callout-warning -->

<!-- #### Question -->

<!-- As we show in Section 13 of this chapter, a random forest will partition the importance of variables across collinear predictors, while a linear regression model (`lm()`) can identify which predictor is causally affecting the response (at least in theory, if all confounders are controlled). What about a boosted regression tree or an artificial neural network? Take the random forest example and add a boosted regression tree (easier, you can use e.g. <https://rdrr.io/cran/xgboost/man/xgb.importance.html>) or an artificial neural network and see if they are better than the random forest at identifying causal predictors. -->

<!-- `r hide("Click here to see the solution")` -->

<!-- ```{r chunk_chapter6_task_1, message=FALSE, warning=FALSE} -->

<!-- library(xgboost) -->

<!-- set.seed(1234) -->

<!-- data = as.data.frame(EcoData::wine) -->

<!-- submission = data[which(is.na(data$quality)), -which(colnames(data) == "quality")] -->

<!-- data = data[complete.cases(data), ] # Removes sumbmission data as well. -->

<!-- data_xg = xgb.DMatrix( -->

<!--   data = as.matrix(data[,which(names(data) != "quality")]), -->

<!--   label = data$quality -->

<!-- ) -->

<!-- brt = xgboost(data_xg, nrounds = 24) -->

<!-- pred = round(predict(brt, newdata = data_xg)) # Only integers are allowed. -->

<!-- table(pred, data$quality) -->

<!-- (accuracy = mean(pred == data$quality)) # Fits very well (on the training data...) -->

<!-- # For submission: -->

<!-- #write.csv(round(predict(rf, submission)), file = "wine_RF.csv") -->

<!-- # Look at variable importance: -->

<!-- xgboost::xgb.importance(model = brt) -->

<!-- ``` -->

<!-- Every method yields slightly different results, but the main ingredient is alcohol (and sulphates). -->

<!-- `r unhide()` -->

<!-- ::: -->

<!-- ::: callout-warning -->

<!-- #### Bonus Task -->

<!-- If you're done with the previous tasks and have still time and appetite, improve the submissions for our competition, in particular for the Wine data set. Possible ideas: -->

<!-- -   Use MLR framework (section @sec-mlr). -->

<!-- -   Try Transfer learning (section @sec-transfer). The winner from last years used transfer learning to win the flower competition -->

<!-- -   Search on kaggle for more ideas / try to copy the ideas. This was the winner two years ago. -->

<!-- A minimal example for the (unbalanced!) Wine data set: -->

<!-- `r hide("Click here to see the solution")` -->

<!-- ```{r chunk_chapter6_task_2, message=FALSE, warning=FALSE,eval=FALSE} -->

<!-- library(tensorflow) -->

<!-- library(keras) -->

<!-- set_random_seed(123L, disable_gpu = FALSE) # Already sets R's random seed. -->

<!-- readin = function(percentageTest = 0.2, aggregate = 0){ -->

<!--     # Parameter "aggregate" packs the classes with very low abundances into one. -->

<!--     # If "aggregate" equals to NA, NaN, Null, 0 or FALSE, no aggregation is performed. -->

<!--     # Else, the given number is the boundary. -->

<!--     # Every class with less elements than the boundary is aggregated into one. -->

<!--     # WARNING: These classes cannot be distinguished from then on! -->

<!--     # Using the predictions for submission needs further processing! -->

<!--     # Just for random selection of features, independent of the amount of function calls. -->

<!--     set.seed(12345) -->

<!--     train = as.data.frame(EcoData::wine) -->

<!--     indicesTrain = which(!is.na(train$quality)) -->

<!--     labelsTrain = train$quality[indicesTrain] -->

<!--     labelsTrain = labelsTrain - min(labelsTrain)  # Start at 0 (for softmax). -->

<!--     train = train[, -which(colnames(train) == "quality")] -->

<!--     if(!is.na(aggregate) & aggregate){ -->

<!--         indices = names(table(labelsTrain)[ -->

<!--             table(labelsTrain) < aggregate & table(labelsTrain) > 0 -->

<!--         ]) -->

<!--         if(length(indices)){ -->

<!--             labelsTrain[labelsTrain %in% indices] = -1 -->

<!--             labelsTrain = as.factor(labelsTrain) -->

<!--             levels(labelsTrain) = 1:length(levels(labelsTrain)) - 1 -->

<!--             labelsTrain = as.integer(labelsTrain) -->

<!--         } -->

<!--     } -->

<!--     # Impute missing values (before any splitting, to get the highest power): -->

<!--     train = missRanger::missRanger( -->

<!--         data = train, -->

<!--         maxiter = 10L, -->

<!--         seed = 123, -->

<!--         num.trees = 200L -->

<!--     ) -->

<!--     # Separate submission data (mind scaling!): -->

<!--     submission = scale(train[-indicesTrain,]) -->

<!--     train = scale(train[indicesTrain,]) -->

<!--     # Very asymmetric training data: -->

<!--     cat(paste0("Size of training set: ", length(labelsTrain), "\n")) -->

<!--     print(table(labelsTrain)) -->

<!--     if(percentageTest == 0){ -->

<!--       return(list( -->

<!--         "labelsTrain" = labelsTrain, -->

<!--         "labelsTest" = list(), -->

<!--         "train" = train, -->

<!--         "test" = list(), -->

<!--         "submission" = submission -->

<!--       )) -->

<!--     } -->

<!--     # Split into training and test set: -->

<!--     len = nrow(train) -->

<!--     indicesTest = sample(x = 1:len, size = percentageTest * len, replace = FALSE) -->

<!--     test = as.data.frame(train[indicesTest,]) -->

<!--     labelsTest = labelsTrain[indicesTest] -->

<!--     train = as.data.frame(train[-indicesTest,]) -->

<!--     labelsTrain = labelsTrain[-indicesTest] -->

<!--     return(list( -->

<!--         "labelsTrain" = labelsTrain, -->

<!--         "labelsTest" = labelsTest, -->

<!--         "train" = train, -->

<!--         "test" = test, -->

<!--         "submission" = submission -->

<!--     )) -->

<!-- } -->

<!-- retVal = readin(aggregate = 0) -->

<!-- labelsTrain = retVal[["labelsTrain"]] -->

<!-- labelsTest = retVal[["labelsTest"]] -->

<!-- train = retVal[["train"]] -->

<!-- test = retVal[["test"]] -->

<!-- submission = retVal[["submission"]] -->

<!-- rm(retVal) -->

<!-- classNumber = length(table(labelsTrain)) -->

<!-- model = keras_model_sequential() -->

<!-- model %>% -->

<!--     layer_dense(units = 200L, activation = "leaky_relu", -->

<!--     kernel_regularizer = regularizer_l2(0.00035), -->

<!--     input_shape = ncol(train)) %>% -->

<!--     layer_dropout(0.45) %>% -->

<!--     layer_dense(units = 100L, activation = "relu", -->

<!--     bias_regularizer = regularizer_l1_l2(0.5)) %>% -->

<!--     layer_dropout(0.2) %>% -->

<!--     layer_dense(units = 100L, activation = "leaky_relu", -->

<!--     kernel_regularizer = regularizer_l2(0.00035), -->

<!--     bias_regularizer = regularizer_l1_l2(0.1)) %>% -->

<!--     layer_dropout(0.25) %>% -->

<!--     layer_dense(units = 50L, activation = "gelu") %>% -->

<!--     layer_dense(units = 25L, activation = "elu") %>% -->

<!--     layer_dropout(0.35) %>% -->

<!--     # We need probabilities. So we use the softmax function. -->

<!--     # Remember, the labels MUST start at 0! -->

<!--     layer_dense(units = classNumber, activation = "softmax") -->

<!-- model %>% -->

<!--     keras::compile(loss = loss_binary_crossentropy, -->

<!--                    optimizer = optimizer_adamax(learning_rate = 0.015)) -->

<!-- model_history =  -->

<!--     model %>% # Mind the matrix property (no data.frame)! -->

<!--         fit(x = as.matrix(train), y = k_one_hot(labelsTrain, classNumber), -->

<!--             epochs = 80L, batch = 12L, shuffle = TRUE) -->

<!-- plot(model_history) -->

<!-- # Accuracy on training set (!) -->

<!-- pred = predict(model, as.matrix(train)) %>% apply(1, which.max) - 1 -->

<!-- Metrics::accuracy(pred, labelsTrain) -->

<!-- table(pred, labelsTrain) -->

<!-- # Accuracy on test set -->

<!-- pred = predict(model, as.matrix(test)) %>% apply(1, which.max) - 1 -->

<!-- Metrics::accuracy(pred, labelsTest) -->

<!-- table(pred, labelsTest) -->

<!-- ``` -->

<!-- Recognize overfitting of your model selection strategy by changing the seed few times (while keeping the model constant) and increase the percentage of test data. Furthermore, consider fitting a random forest for good quality as well. -->

<!-- For the final predictions, we use the whole data set without holdouts: -->

<!-- ```{r chunk_chapter6_task_3, message=FALSE, warning=FALSE, eval=FALSE} -->

<!-- library(tensorflow) -->

<!-- library(keras) -->

<!-- set_random_seed(321L, disable_gpu = FALSE) # Already sets R's random seed. -->

<!-- retVal = readin(percentageTest = 0, aggregate = 0) -->

<!-- labelsTrain = retVal[["labelsTrain"]] -->

<!-- labelsTest = retVal[["labelsTest"]] -->

<!-- train = retVal[["train"]] -->

<!-- test = retVal[["test"]] -->

<!-- submission = retVal[["submission"]] -->

<!-- rm(retVal) -->

<!-- classNumber = length(table(labelsTrain)) -->

<!-- model = keras_model_sequential() -->

<!-- model %>% -->

<!--     layer_dense(units = 200L, activation = "leaky_relu", -->

<!--     kernel_regularizer = regularizer_l2(0.00035), -->

<!--     input_shape = ncol(train)) %>% -->

<!--     layer_dropout(0.45) %>% -->

<!--     layer_dense(units = 100L, activation = "relu", -->

<!--     bias_regularizer = regularizer_l1_l2(0.5)) %>% -->

<!--     layer_dropout(0.2) %>% -->

<!--     layer_dense(units = 100L, activation = "leaky_relu", -->

<!--     kernel_regularizer = regularizer_l2(0.00035), -->

<!--     bias_regularizer = regularizer_l1_l2(0.1)) %>% -->

<!--     layer_dropout(0.25) %>% -->

<!--     layer_dense(units = 50L, activation = "gelu") %>% -->

<!--     layer_dense(units = 25L, activation = "elu") %>% -->

<!--     layer_dropout(0.35) %>% -->

<!--     # We need probabilities. So we use the softmax function. -->

<!--     # Remember, the labels MUST start at 0! -->

<!--     layer_dense(units = classNumber, activation = "softmax") -->

<!-- model %>% -->

<!--     keras::compile(loss = loss_binary_crossentropy, -->

<!--                    optimizer = optimizer_adamax(learning_rate = 0.015)) -->

<!-- model_history =  -->

<!--     model %>% # Mind the matrix property (no data.frame)! -->

<!--         fit(x = as.matrix(train), y = k_one_hot(labelsTrain, classNumber), -->

<!--             epochs = 80L, batch = 12L, shuffle = TRUE) -->

<!-- plot(model_history) -->

<!-- # Accuracy on training set (!) -->

<!-- pred = predict(model, as.matrix(train)) %>% apply(1, which.max) - 1 -->

<!-- Metrics::accuracy(pred, labelsTrain) -->

<!-- table(pred, labelsTrain) -->

<!-- # Reverse subtraction (for start at 0) and create submission file. -->

<!-- write.csv(pred + min(as.data.frame(EcoData::wine)$quality, na.rm = TRUE), -->

<!--           file = "wine_NN.csv") -->

<!-- ``` -->

<!-- `r unhide()` -->

<!-- ::: -->
