# Interpretation and causality with machine learning

## Explainable AI

The goal of explainable AI (xAI, aka interpretable machine learning) is to explain WHY a fitted ML models makes certain predictions. A typical example is to understand  how important different variables are for predictions. There incentives to do so range from a better technical understanding of the models over understanding which data is important to improve predictions to questions of fairness and discrimination (e.g. to understand if an algorithm uses skin color to make a decision).

### A practical example

In this lecture we will work with another famous dataset, the Boston housing dataset:

We will fit a random forest and use the iml pkg for xAI, see ![](https://christophm.github.io/interpretable-ml-book/)

```{r}
set.seed(123)
library("iml")
library("randomForest")
data("Boston", package = "MASS")
rf = randomForest(medv ~ ., data = Boston, ntree = 50)
```

xAI packages are written generic, i.e. they can handle almost all ML models. 
When we want to use them, we first have to create a Predictor object, that holds the model and the data. The iml package uses R6 classes, that means new objects can be created by calling Predictor$new(). (do not worry if you do not know what R6 classes are, just use the command)

```{r}
X = Boston[which(names(Boston) != "medv")]
predictor = Predictor$new(rf, data = X, y = Boston$medv)
```

### Feature Importance
Feature importance, should not be mistaken with the RF variable importance. It tells us how important the individual variables are for predictions and can be calculated for all ML models and is based on a permutation approach (have a look at the book):

```{r}
imp = FeatureImp$new(predictor, loss = "mae")
plot(imp)
```

### Partial dependencies

Partial dependencies are similar to allEffects plots for normal regressions, the idea is to visualize "marginal effects" of predictors (with the feature argument we specify the variable we want to visualize):

```{r}
eff = FeatureEffect$new(predictor, feature = "rm", method = "pdp", grid.size = 30)
plot(eff)
```

Partial dependencies can be also plotted for single observations:

```{r}
eff = FeatureEffect$new(predictor, feature = "rm", method = "pdp+ice", grid.size = 30)
plot(eff)
```

One disadvantage of partial dependencies is that they are sensitive to correlated predictors. Accumulated local effects can be used to account for correlation for predictors

### Accumulated local effects 

Accumulated local effects (ALE) are basically partial dependencies plots but try to correct for correlations between predictors
```{r}
ale = FeatureEffect$new(predictor, feature = "rm", method = "ale")
ale$plot()
```

If there is no colinearity, you shouldn't see much difference between partial dependencies and ALE plots.

### Friedmans H-statistic

The H-statistic can be used to find interactions between predictors. However, again, keep in mind that the H-statistic is sensible to correlation between predictors:

```{r}
interact = Interaction$new(predictor, "lstat")
plot(interact)
```

### Global explainer - Simplifying the ML model

Another idea is to simplify the ML model with another simpler model such as a decision tree. We create predictions with the ML model for a lot of different input values and then we fit on these predictions a decision tree, which we can then interpret.

```{r}
library(partykit)
tree = TreeSurrogate$new(predictor, maxdepth = 2)
plot(tree)
```

### Local explainer - LIME explaining single instances (observations)

The global approach is to simplify the entire ML-black-box model via a simpler model, which is then interpretable. 

However, sometimes we are only interested in understanding how single observations/predictions are generated. The lime approach explores the feature space around one observations and based on this local spare fits then a simpler model (e.g. a linear model):

```{r}
library(glmnet)
lime.explain = LocalModel$new(predictor, x.interest = X[1,])
lime.explain$results
plot(lime.explain)
```


### Local explainer - Shapley 

The Shapley method computes the so called Shapley value, feature contributions for single predictions, and is based on an approach from cooperative game theory. The idea is that each feature value of the instance is a "player" in a game, where the prediction is the reward. The Shapley value tells us how to fairly distribute the award among the feature. 

```{r}
shapley = Shapley$new(predictor, x.interest = X[1,])
shapley$plot()
```

## Causal machine learning

In general, xAI != causality. xAI methods measure which variables are used by the algorithm for predictions, or how much variables improve predictions. However, it is very often possible that a variable that does not cause something can predict something. So you can not interpret predictive power with causality.

In statistical courses (in particular course: advanced biostatistics), we discuss the issue of causality at length. Here, we don't want to go into the details, but again, you should in general resist to interpret indicators of importance in xAI as causal effects. They tell you something about what's going on in the algorithm, not about what's going on in reality.

We will illustrate the problem at an example for the variable importance in the RF algorithm. The  RF variable importance will split importance values for colinear variables evenly. Thus, colinearty introduces importance even if colinearity is low enough so that variables would be separable and would be correctly separated by an lm / ANOVA.

We simulate a dataset with 2 predictors that are strongly correlated, but only one of them has an effect on the response. 
```{r}
# simulation parameters
n = 1000
col = 0.7
# create collinear predictors
x1 = runif(n)
x2 = col * x1 + (1-col) * runif(n)
# response is only influenced by x1
y = x1 + rnorm(n)
```
lm / anova correctly identify x1 as causal variable
```{r}
anova(lm(y ~ x1 + x2))
```

Fit RF and show variable importance
```{r}
fit <- randomForest(y ~ x1 + x2, importance=TRUE)
varImpPlot(fit)
```
Variable importance is now split nearly evenly.

Task: understand why this is - remember: 

- How the random forest works - variables are randomly hidden from the regression tree when the trees for the forest are built
- Remember that as x1 ~ x2, we can use x2 as a replacement for x1
- Remember that the variable importance measures the average contributions of the different variables in the trees of the forest

### Causal inference on static data 

Pearl


Automatic discovery 

### Causal inference on dynamic data 

Granger causality


### Transferring these ideas to ML

N



