--- 
title: "A Minimal Book Example"
author: "Maximilian Pichler and Florian Hartig"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---
--- 
title: "A Minimal Book Example"
author: "Maximilian Pichler and Florian Hartig"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Prerequisites

R+Rstudio

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->



# Introduction to Machine Learning {#introduction}
In this lesson, we introduce the three basic ML tasks: supervised regression and classification, and unsupervised learning. 

In ML, we distinguish 3 basic learning paradigms:
- Supervised learning
- Unsupervised learning
- Reinforcement learning
We will speak about reinforcement learning at the end of the course. Now, we want to look at examples of supervised and unsupervised learning. Before you start with the code, here a video to remind you of what we talked about in the class:


```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe width="560" height="315" 
  src="https://www.youtube.com/embed/1AVrWvRvfxs"
  frameborder="0" allow="accelerometer; autoplay; encrypted-media;
  gyroscope; picture-in-picture" allowfullscreen>
  </iframe>'
)
```


## Supervised learning: regression and classification
Two two main subbranches of supervised learning are regression and classification. Here a video that explains again the difference 

```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe width="560" height="315" 
  src="https://www.youtube.com/embed/i04Pfrb71vk"
  frameborder="0" allow="accelerometer; autoplay; encrypted-media;
  gyroscope; picture-in-picture" allowfullscreen>
  </iframe>'
)
```


#### Supervised regression using Random Forest

The random forest (RF) algorithm is possibly the most widely used ML algorithm and can be used for regression and classification. We will talk more about the algorithm on Day 2. Here an example of a regression: 

Visualization of the data:
```{r}
plot(iris, col = iris$Species)
```

Fitting the model
```{r}
library(randomForest)
m1 <- randomForest(Sepal.Length ~ ., data = iris)
# str(m1)
# m1$type
# predict(m1)
print(m1)
```

Visualization of the results
```{r}
par(mfrow = c(1,2))
plot(predict(m1), iris$Sepal.Length, xlab = "predicted", ylab = "observed")
abline(0,1)
varImpPlot(m1)
```

This is a nice visualization of the RF structure, but requires to load a package from GitHub

```{r}
# devtools::install_github('araastat/reprtree')
reprtree:::plot.getTree(m1, iris)
```

#### Supervised classification using Random Forest
Fitting the model:
```{r}
set.seed(123)
m1 <- randomForest(Species ~ ., data = iris)
# str(m1)
# m1$type
# predict(m1)
print(m1)
```
Visualizing the fitted model:

```{r}
par(mfrow = c(1,2))
reprtree:::plot.getTree(m1, iris)
```

Visualizing results ecologically:
```{r}
oldpar <- par(mfrow = c(1,2))
plot(iris$Petal.Width, iris$Petal.Length, col = iris$Species, main = "observed")
plot(iris$Petal.Width, iris$Petal.Length, col = predict(m1), main = "predicted")
```

```{r,echo=FALSE}
par(oldpar)
```


Confusion matrix:
```{r}
table(predict(m1),iris$Species)
```


### Unsupervised learning

In unsupervised learning, we basically want to identify patterns in data without having any guidance (supervision) about what the correct patterns / classes are.

It is all much easier with a practical example. Consider our iris dataset.

- Here, we have observations of different species
- Together with their flower traits
Imagine we didn't know what species are. This is basically the situation in which people in the antique would have been. There is no book to look up species. You just noted that there seem to be some kind of plants that have different flowers than another, so you decide to call them by a different name. This kind of process is what unsupervised learning does.

#### k-means clustering
An example for an unsupervised learning algorithm is k-means clustering, one of the simplest and popular unsupervised machine learning algorithms.

A cluster refers to a collection of data points aggregated together because of certain similarities. In the algorithm, you’ll define a target number k, which refers to the number of centroids you need in the dataset. A centroid is the imaginary or real location representing the center of the cluster.

Every data point is allocated to each of the clusters through reducing the in-cluster sum of squares. In other words, the K-means algorithm identifies k number of centroids, and then allocates every data point to the nearest cluster, while keeping the centroids as small as possible. The ‘means’ in the K-means refers to averaging of the data; that is, finding the centroid.

```{r}
sIris = scale(iris[,1:4])
model<- kmeans(sIris,3) # aplly k-means algorithm with no. of centroids(k)=3
model
```

Visualizing the results:

```{r}
par(mfrow = c(1,2))
plot(Petal.Length~Petal.Width, data = sIris, col = model$cluster, main = "Predicted clusters")
plot(Petal.Length~Petal.Width, data = sIris, col = iris$Species, main = "True species")
```

Confusion matrix:
```{r}
table(model$cluster,iris$Species)
```



## Introduction to Tensorflow {#tf}


```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe width="560" height="315" 
  src="https://www.youtube.com/embed/MotG3XI2qSs"
  frameborder="0" allow="accelerometer; autoplay; encrypted-media;
  gyroscope; picture-in-picture" allowfullscreen>
  </iframe>'
)
```


- TF is a math library which is highly optimized for neural networks
- If a GPU is available, computations can be easily run on the GPU but even on the CPU is TF still very fast
- The "backend" (i.e. all the functions and all computations) are written in C++ and CUDA (CUDA is a programming language for the GPU)
- The interface (the part of TF that we use) is written in python and is also available in R, which means, we can write the code in R/Python but it will be executed by the (compiled) C++ backend. 

### Tensorflow data containers
TF has two data containers (structures):
- constant (tf$constant) :creates a constant (immutable) value in the computation graph
- variable (tf$Variable): creates a mutable value in the computation graph (used as parameter/weight in models)

```{r}
library(tensorflow)
# Don't worry about weird messages. TF supports additional optimizations
exists("tf")
```
Don't worry about weird messages (they will only appear once at the start of the session).


```{r}
a = tf$constant(5)
b = tf$constant(10)
print(a)
print(b)
c = tf$add(a, b)
print(c)
tf$print(c)
```

Normal R methods such as print() are provided by the R package "tensorflow". 

The tensorflow library (created by the RStudio team) built R methods for all common operations:

```{r}
`+.tensorflow.tensor` = function(a, b) return(tf$add(a,b))
tf$print(a+b)
```

Their operators also transfrom automatically R numbers into constant tensors when attempting to add a tensor to a R number:
```{r}
d = c + 5  # 5 is automatically converted to a tensor
print(d)
```

TF container are objects, which means that they are not just simple variables of type numeric (class(5)), and they have methods
For instance, there is a method to transform the tensor object back to a R object:

```{r}
class(d)
class(d$numpy())
```

### Tensorflow data types - good practise with R-TF
R uses dynamic typing, which means you can assign to a variable a number, character, function or whatever, and R infers the type automatically.
In other languages you have to state explicitly the type, e.g. in C: int a = 5; float a = 5.0; char a = "a";
While TF tries to infer dynamically the type, often you must state it explicitly.
Common important types: 
- float32 (floating point number with 32bits, "single precision")
- float64 ("double precision")
- int8 (integer with 8bits)
Why does TF support float32 numbers when most cpus today can handle easily 64bit numbers? 
Many GPUs (e.g. the NVIDIA geforces) can handle only up to 32bit numbers! (you do not need high precision in graphical modeling)

```{r}
r_matrix = matrix(runif(10*10), 10,10)
m = tf$constant(r_matrix, dtype = "float64") 
b = tf$constant(2.0, dtype = "float64")
c = m / b
```

Instead of a string, you can also provide a tf$float64 object
```{r}
m = tf$constant(r_matrix, dtype = tf$float32)
b = tf$constant(2.0, dtype = tf$float64)
c = m / b # doesn't work! we try to divide float32/float64
```

Tensorflow arguments often require exact/explicit data types:
TF expects for arguments integers, however, R needs a "L" after an integer to tell the R interpreter that it should be treated as an integer:

```{r}
is.integer(5)
is.integer(5L)
matrix(t(r_matrix), 5, 20, byrow = TRUE)
tf$reshape(r_matrix, shape = c(5, 20))$numpy()
tf$reshape(r_matrix, shape = c(5L, 20L))$numpy()
```

Note: skipping the "L" is one of the most common errors when using R-TF!


<!--chapter:end:01-intro.Rmd-->

# Introduction to Tensorflow {#tf}


```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe width="560" height="315" 
  src="https://www.youtube.com/embed/MotG3XI2qSs"
  frameborder="0" allow="accelerometer; autoplay; encrypted-media;
  gyroscope; picture-in-picture" allowfullscreen>
  </iframe>'
)
```


- TF is a math library which is highly optimized for neural networks
- If a GPU is available, computations can be easily run on the GPU but even on the CPU is TF still very fast
- The "backend" (i.e. all the functions and all computations) are written in C++ and CUDA (CUDA is a programming language for the GPU)
- The interface (the part of TF that we use) is written in python and is also available in R, which means, we can write the code in R/Python but it will be executed by the (compiled) C++ backend. 

<!--chapter:end:02-tensorflow.Rmd-->

