--- 
title: "Machine Learning and AI in TensorFlow and R"
author: "Maximilian Pichler and Florian Hartig"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:
    highlight: kate
documentclass: book
bibliography: ["packages.bib", "literature.bib"]
biblio-style: "apalike"
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Machine Learning and AI in TensorFlow and R"
---

# Prerequisites

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

knitr::opts_chunk$set(cache=TRUE)
``` 

**R system**

Make sure you have a recent version of R (>=3.6, ideally >=4.0) and RStudio on your computers. 

**TensorFlow and Keras**

If you want to run the code on your own computers, you also will need to install TensorFlow / Keras for R. For this, the following should work for most people:

Run in R: 
```{r, eval=FALSE}
install.packages("keras", dependencies = T)
keras::install_keras()
```

This should work on most computers, in particular if all software is recent. Sometimes, however, things don't work well, especially the python distribution often makes problems. If the installation does not work for you, we can look at it together. Also, we will provide some virtual machines in case your computers / laptops are too old or you don't manage to install TensorFlow.

**Torch for R**

We may also use Torch for R. This is an R frontend for the popular PyTorch framework. To install Torch, type in R:

```{r,eval=FALSE}
install.packages("torch")
library(torch)
```

**EcoData**

Finally, we may sometimes use datasets from the EcoData package. To install the package, run:
```{r,eval=FALSE}
devtools::install_github(repo = "florianhartig/EcoData", subdir = "EcoData", 
dependencies = TRUE, build_vignettes = TRUE)
```
