--- 
title: "Machine Learning and AI in TensorFlow and R"
author: "Maximilian Pichler and Florian Hartig"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:
    highlight: kate
documentclass: book
bibliography: ["packages.bib", "literature.bib"]
biblio-style: "apalike"
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Machine Learning and AI in TensorFlow and R"
---

# Prerequisites

```{r, include=FALSE}
# Automatically create a bib database for R packages.
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
                 'packages.bib')

knitr::opts_chunk$set(cache = TRUE)
``` 


## R system

Make sure you have a recent version of R (>=3.6, ideally >=4.0) and RStudio on your computers. 


## TensorFlow and Keras

If you want to run the code on your own computers, you also will need to install TensorFlow / Keras for R. For this, the following should work for most people:

Run in R: 

```{r, eval=FALSE}
install.packages("keras", dependencies = T)
keras::install_keras()
```

This should work on most computers, in particular if all software is recent. Sometimes, however, things don't work well, especially the python distribution often makes problems. If the installation does not work for you, we can look at it together. Also, we will provide some virtual machines in case your computers / laptops are too old or you don't manage to install TensorFlow.


## Torch for R

We may also use Torch for R. This is an R frontend for the popular PyTorch framework. To install Torch, type in R:

```{r, eval=FALSE}
install.packages("torch")
library(torch)
```


## EcoData

We may sometimes use data sets from the EcoData package. To install the package, run:

```{r, eval=FALSE}
devtools::install_github(repo = "florianhartig/EcoData", subdir = "EcoData", 
dependencies = TRUE, build_vignettes = TRUE)
```


## Further used libraries

We will make huge use of different libraries. So take a coffee or two (that will take a while...) and install the following libraries:

```{r, eval=FALSE}
install.packages("abind")
install.packages("animation")
install.packages("ape")
install.packages("BiocManager")
BiocManager::install(c("Rgraphviz", "graph", "RBGL"))
install.packages("coro")
install.packages("dbscan")
install.packages("dendextend")
install.packages("devtools")
install.packages("dplyr")
install.packages("e1071")
install.packages("factoextra")
install.packages("fields")
install.packages("forcats")
install.packages("glmnet")
install.packages("gym")
install.packages("kknn")
install.packages("knitr")
install.packages("iml")
install.packages("lavaan")
devtools::install_version("lavaanPlot", version = "0.6.0")
install.packages("lmtest")
install.packages("mclust")
install.packages("missRanger")
install.packages("mlr3")
install.packages("mlr3learners")
install.packages("mlr3measures")
install.packages("mlr3pipelines")
install.packages("mlr3tuning")
install.packages("paradox")
install.packages("partykit")
install.packages("pcalg")
install.packages("piecewiseSEM")
install.packages("purrr")
install.packages("randomForest")
install.packages("ranger")
install.packages("reticulate")
devtools::install_github('araastat/reprtree')
install.packages("rpart")
install.packages("rpart.plot")
install.packages("scales")
install.packages("semPlot")
install.packages("stringr")
install.packages("tfprobability")
install.packages("tidyverse")
install.packages("torchvision")
install.packages("xgboost")
```


## Linux/UNIX systems have to fulfill some further dependencies

**Debian based systems**

For Debian based systems, we need:

```{}
libmagick++-dev
build-essential
gfortran
```

If you are new to installing packages on Debian, type the following:

```{}
sudo apt update && sudo apt install -y libmagick++-dev build-essential gfortran
```


